Performance for the masses: Experiments with A Web Based Architecture to Harness Volunteer Resources for Low Cost Distributed Evolutionary Computation

### 2016-02-02 GECCO/volunteer-clouds-gecco16.tex
Rewriting conclusions
Mainly to get back to 8 pages. Abstract has been quite revised for now, so this closes #96

In this paper we will carry out three experiments, after establishing
the baseline of the performance that can be expected in a JavaScript
based system and its comparison with other languages in
\cite{2016arXiv160101607Manom}. First, several experiments using a
Platform as a Service, and a simple Ajax client that does not
restart. Second, we change a bit the client so that it uses web
workers, that is, it runs in the background and it restarts. Third, we use the same
client but with a harder problem that cannot be solved by a single
client and thus needs the contribution of many volunteers to
proceed. With these three experiments we want mainly to try and model
the behavior of users and how different problems and implementations
have an influence on it.

@@

Other systems, like JET, proposed by Soares et
al. \cite{soares1998get}, also used Java. JET supports
the execution of parallel applications over the Internet and features
a comprehensive statistics collection subsystem, enabling its use for science, %(Mario) Was easier to change "allowing"
since it allows % (Paloma) Use synonyms, like "yield" or "permit" 
the collection of algorithm analytics by the
experiment designer, unlike the other systems mentioned
above. Other proposals contemporary to JET, such as {\em SuperWeb}
\cite{alexandrov1997superweb} are roughly a translation of classical
distribution techniques, with overly complicated architectures that include
brokers for resource discovery. SuperWeb is focused,
% (Paloma) Careful with starting using plural ("Other proposals", "are") and continuig with singular. If you are referring to "SuperWeb", change "It" for "SuperWeb"
however, in solving the
issues involved in browser-based computing and also in making a system
that is able to work across different types of computers, mentioning
SPARCstations and PCs with Windows-95. Interestingly enough, this
paper incorporates a discussion on the economic model, paving the way
for the for-profit distributed computing systems that arose later on. 
On the other hand, the potential for volunteer computing using
browsers was realized
later on \cite{sarmenta-bayanihan} as well as the potential of
user misconduct when having access to the source code \cite{sarmenta-sabotagetolerance}. 
However, these early efforts by
Sarmenta once again used Java and not JavaScript, making them % (Paloma) making them
less widespread.

The main problem with using Java and all the other browser-embedded
technologies is that they are not {\em universal} in the
sense that an extra component, namely, the Java virtual machine or another
virtual environment has to be installed in the browser.
Also these virtual environments have become targets of malicious attacks; 
for instance numerous vulnerabilities have been discovered in 
the proprietary Adobe Flash runtime environment and
video player \cite{ford2009analyzing,watanabe2010new}. The need for
these additional environments has been mitigated with the
adoption of the HTML5 standard \cite{anthes2012html5}.
JavaScript however, has become a set of standards that include the language and its components.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{puts-openshift-4-4.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-4-24.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-7-31.png}
\caption{Number of PUTs per unique IP and fit to a Generalized
  Extreme Value distribution (in a lighter shade or blue). From left to right, experiments
  4/4, 4/24 and 7/31.} 
% Plotted with ../data/plot-zipf-openshift.R
\label{fig:puts:os}
\end{figure}
%
It is also interesting to check the distribution of the experiment
duration, shown in Figure \ref{fig:zipf:os} and which roughly follows
a Zipf's law, with similar distribution along all three runs. The 4/24
run is the most complete and shows a S-shape, which implies an
accumulation of experiments taking similar time and around 100
seconds. The most interesting part is the {\em tail}, which shows how
many experiments took a desirable amount of time, of the order of
10 seconds, and which appears in all three graphs. As it can be seen,
it sharply drops implying there are 
just a few of them, and with diminishing probability as time
decreases. This exponential distribution also appears in the
distribution of HTTP PUTs, equivalent to the number of
generations divided by 100, contributed by every user, which is shown
in Figure \ref{fig:puts:os}. These results show a Zipf-like behavior,
so we have fitted it to the Generalized Extreme Value distribution,
with the resulting parameters shown in Table \ref{tab:puts:os}.
%
\begin{table*}
\caption{Summary of fit to Generalized Extreme Value distribution of
  the number of PUTs per unique IP. \label{tab:puts:os}}
\begin{center}
\begin{tabular}{l|ccc}
\hline
Date  & Location $\mu$ & Scale $\sigma$ & Shape $\xi$ \\
\hline
4/4 &  8.541 $\pm$ 1.0926  &    12.442 $\pm$ 1.7302 &  1.388 $\pm$
0.1377 \\
4/24 & 6.148 $\pm$ 0.3782 & 7.354 $\pm$ 0.5105 & 1.090 $\pm$  0.0697  \\
7/31 & 11.645 $\pm$ 1.475 & 16.365 $\pm$ 2.201 &  1.265 $\pm$ 0.132   \\
\hline
\end{tabular}
\end{center}
\end{table*}
%
This distribution was originally proposed to fit extreme values
\cite{resnick2013extreme} and contains, as a special case, the inverse
Weibull distribution which was fitted to volunteer computing
frameworks such as SETI@home \cite{javadi2009mining}. We obviously do
not pretend to compare our framework in scale or complexity with it, but
to point out that the behavior of volunteer computing nodes follows a
certain pattern, found in SETI@home, and which also appears in our framework. This
distribution is governed by three parameters, the usual location $\mu$
which is related to where it has its {\em center} and a scale $\sigma$,
related to the size, but also a third shape $\xi$ parameter that is
related to its skewness, that is, how skewed it is around the central
location. Positive parameters indicate that the distribution {\em
  leans} towards the origin, and negative ones towards the other extreme
value. In this case, Table \ref{tab:puts:os} shows $\xi$ values
greater than one and between one and 1.4, which indicates that the
three experiments share this origin-leaning pattern, with many users
donating a few cycles and just a few donating extreme values. Random
distributions with these parameters have been plotted in red in
Figure \ref{fig:puts:os}, indicating that the fit is good enough. The
predictive value of these fits, however, is limited, over all taking
into account the low correlation between successive events shown
above.

@@

The experiments were run in the same way as before, using several
 announcements in Twitter throughout several days by the end of July
2015. Eventually, more than one thousand experiments were completed in
a matter of days. A summary of results is shown in Table
\ref{tab:summary:ww}, comparing it with the aggregate of the results
obtained with the initial version of {\sf NodIO}. These results are
remarkably different, being similar only in the median and maximum
number of IPs, although in this case it would be combination
IP-worker, as we considered every worker as a unique IP. The median
number of seconds is almost 80\% better than the previous version, and
the median number of PUTs is three times better; this causes that most
of experiments take less time than one of the baselines, and one third
less than the second baseline. Thus, the conclusion in this case is
that we can obtain, through volunteer computation, a result that most 
of the times is better than we would by using a similar, non-parallel,
desktop setup, which makes this system suitable for massively
distributed evolutionary algorithms.
%
\begin{table*}
\caption{Summary of fit to GEV and Weibull distribution of
  the number of PUTs per worker. \label{tab:puts:ww}}
\begin{center}
\begin{tabular}{cccc}
\hline
Distribution & Location $\mu$ & Scale $\sigma$ & Shape $\xi$ \\
\hline
GEV & 18.275 $\pm$ 0.83719  &  25.962  $\pm$ 1.22015 & 1.242   $\pm$
0.04746 \\
Weibull & ND & 75.11457519 $\pm$ 2.97470251  & 0.69714410 $\pm$ 0.01383976 \\
\hline
\end{tabular}
\end{center}
\end{table*}
%

@@

%---------------------------------------------------------------
\section{Conclusion}
\label{sec:conclusion}

@@

In general, linking and finding correlations between user choices and
performance is an interesting avenue to explore in the future. Even if
the three previous experiments were published in a similar way, one
obtained up to 5 times more total cycles  than the one with the least
number of cycles. It is also essential to obtain volunteers as
simultaneously as possible, so it is possible that the features of the
social network in terms of real-time use will also play a big
role. Even as it is difficult to create controlled experiments in this
area, it is an interesting challenge to face in the future. % (Paloma) Wrong collocation

Although we think that, in general, the results presented in this
paper are independent of the problem chosen, it is true that even if
it is a difficult problem for evolutionary algorithms, it takes a few
seconds with the right settings to solve. This makes it difficult also
for this kind of asynchronous framework, but it would be interesting
to check the performance for other computationally demanding
tasks, especially ones that cover the middle ground between a problem
finishing in seconds or minutes (Trap) and weeks (Rastrigin F15). 
Besides, the two versions of the framework include as an
algorithmic variant using random population size. We do not think this
has had a big influence on the results and in fact this is not noticed
for F15, which is the second version tested, but it would be interesting to
measure exactly what this influence has been. 
% In general there are
% many issues with the evolutionary algorithm implementation itself,
% including using different, or adaptive, policies for inserting and
% sending individuals to the pool as was done in \cite{araujo2008mam},
% using different policies for population initialization, and also the
% incorporation of high-speed local resources to the pool to check what
% would be the real influence of the volunteer pool to the final
% performance.

@@

Finally, the implementation needs some refinement in terms of
programming and also ease of use. Tools such as Yeoman might be used % ¿añadir una cita o la URL http://yeoman.io/ a pie de página ?
to create a generator in which the user just has to create a fitness
function, with the rest of the framework wrapped around
automatically.
%  All this, as well as the data used for this paper and the paper
% itself, has been published with a free license in GitHub at
% \url{https://github.com/JJ/modeling-volunteer-computing}.

>>>>

In this paper we will carry out three experiments, after establishing
the baseline of the performance that can be expected in a JavaScript
based system and its comparison with other languages in
\cite{2016arXiv160101607Manom}. First, several experiments using a
Platform as a Service, and a simple Ajax client that does not
restart. Second, we change a bit the client so that it uses web
workers, that is, it runs in the background and it restarts. Third, we use the same
client but with a harder problem that cannot be solved by a single
volunteer and thus needs the contribution of many volunteers to
proceed. With these three experiments we want mainly to try and model
the behavior of users and how different problems and implementations
have an influence on it.

@@

Other systems, like JET, proposed by Soares et
al. \cite{soares1998get}, also used Java. JET supports
the execution of parallel applications over the Internet and features
a comprehensive statistics collection subsystem, enabling its use for science, %(Mario) Was easier to change "allowing"
since it allows % (Paloma) Use synonyms, like "yield" or "permit" 
the collection of algorithm analytics by the
experiment designer, unlike the other systems mentioned
above. Other proposals contemporary to JET, such as {\em SuperWeb}
\cite{alexandrov1997superweb} are roughly a translation of classical
distribution techniques, with overly complicated architectures that include
brokers for resource discovery. SuperWeb is focused,
% (Paloma) Careful with starting using plural ("Other proposals", "are") and continuig with singular. If you are referring to "SuperWeb", change "It" for "SuperWeb"
however, in solving the
issues involved in browser-based computing and also in making a system
that is able to work across different types of computers. Interestingly enough, this
paper incorporates a discussion on the economic model, paving the way
for the for-profit distributed computing systems that arose later on. 
On the other hand, the potential for volunteer computing using
browsers was realized
later on \cite{sarmenta-bayanihan} as well as the potential of
user misconduct when having access to the source code \cite{sarmenta-sabotagetolerance}. 
However, these early efforts by
Sarmenta once again used Java and not JavaScript, making them % (Paloma) making them
less widespread.

The main problem with using Java and all the other browser-embedded
technologies is that they are not {\em universal} in the
sense that an extra component, namely, the Java virtual machine or another
virtual environment has to be installed in the browser.
% Also these virtual environments have become targets of malicious attacks; 
% for instance numerous vulnerabilities have been discovered in 
% the proprietary Adobe Flash runtime environment and
% video player \cite{ford2009analyzing,watanabe2010new}. 
The need for
these additional environments has been mitigated with the
adoption of the HTML5 standard \cite{anthes2012html5}, which is
tightly integrated with JavaScript.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{puts-openshift-4-4.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-4-24.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-7-31.png}
\caption{Number of PUTs per unique IP and fit to a Generalized
  Extreme Value distribution (in a lighter shade or blue). From left to right, experiments
  4/4, 4/24 and 7/31.} 
% Plotted with ../data/plot-zipf-openshift.R
\label{fig:puts:os}
\end{figure}
%
It is also interesting to check the distribution of the experiment
duration, shown in Figure \ref{fig:zipf:os} and which roughly follows
a Zipf's law, with similar distribution along all three runs. The 4/24
run is the most complete and shows a S-shape, which implies an
accumulation of experiments taking similar time and around 100
seconds. The most interesting part is the {\em tail}, which shows how
many experiments took a desirable amount of time, of the order of
10 seconds, and which appears in all three graphs. As it can be seen,
it sharply drops implying there are 
just a few of them, and with diminishing probability as time
decreases. This exponential distribution also appears in the
distribution of HTTP PUTs, equivalent to the number of
generations divided by 100, contributed by every user, which is shown
in Figure \ref{fig:puts:os}. These results show a Zipf-like behavior,
so we have fitted it to the Generalized Extreme Value distribution,
with the resulting parameters shown in Table \ref{tab:puts:os}.
%
\begin{table}
\caption{Summary of fit to Generalized Extreme Value distribution of
  the number of PUTs per unique IP. \label{tab:puts:os}}
\begin{center}
\begin{tabular}{l|ccc}
\hline
Date  & Location $\mu$ & Scale $\sigma$ & Shape $\xi$ \\
\hline
4/4 &  8.541 $\pm$ 1.0926  &    12.442 $\pm$ 1.7302 &  1.388 $\pm$
0.1377 \\
4/24 & 6.148 $\pm$ 0.3782 & 7.354 $\pm$ 0.5105 & 1.090 $\pm$  0.0697  \\
7/31 & 11.645 $\pm$ 1.475 & 16.365 $\pm$ 2.201 &  1.265 $\pm$ 0.132   \\
\hline
\end{tabular}
\end{center}
\end{table}
%
This distribution was originally proposed to fit extreme values
\cite{resnick2013extreme} and contains, as a special case, the inverse
Weibull distribution which was fitted to volunteer computing
frameworks such as SETI@home \cite{javadi2009mining}. We obviously do
not pretend to compare our framework in scale or complexity with it, but
to point out that the behavior of volunteer computing nodes follows a
certain pattern, found in SETI@home, and which also appears in our framework. This
distribution is governed by three parameters, the usual location $\mu$
which is related to where it has its {\em center} and a scale $\sigma$,
related to the size, but also a third shape $\xi$ parameter that is
related to its skewness, that is, how skewed it is around the central
location. Positive parameters indicate that the distribution {\em
  leans} towards the origin, and negative ones towards the other extreme
value. In this case, Table \ref{tab:puts:os} shows $\xi$ values
greater than one and between one and 1.4, which indicates that the
three experiments share this origin-leaning pattern, with many users
donating a few cycles and just a few donating extreme values. Random
distributions with these parameters have been plotted in red in
Figure \ref{fig:puts:os}, indicating that the fit is good enough. The
predictive value of these fits, however, is limited, over all taking
into account the low correlation between successive events shown
above.

@@

The experiments were run in the same way as before, using several
 announcements in Twitter throughout several days by the end of July
2015. Eventually, more than one thousand experiments were completed in
a matter of days. A summary of results is shown in Table
\ref{tab:summary:ww}, comparing it with the aggregate of the results
obtained with the initial version of {\sf NodIO}. These results are
remarkably different, being similar only in the median and maximum
number of IPs, although in this case it would be combination
IP-worker, as we considered every worker as a unique IP. The median
number of seconds is almost 80\% better than the previous version, and
the median number of PUTs is three times better; this causes that most
of experiments take less time than one of the baselines, and one third
less than the second baseline. Thus, the conclusion in this case is
that we can obtain, through volunteer computation, a result that most 
of the times is better than we would by using a similar, non-parallel,
desktop setup, which makes this system suitable for massively
distributed evolutionary algorithms.
%
\begin{table}
\caption{Summary of fit to GEV and Weibull distribution of
  the number of PUTs per worker. \label{tab:puts:ww}}
\begin{center}
\begin{tabular}{cccc}
\hline
Distribution & Location $\mu$ & Scale $\sigma$ & Shape $\xi$ \\
\hline
GEV & 18.275 $\pm$ 0.837  &  25.96  $\pm$ 1.22 & 1.242   $\pm$ 0.047 \\
Weibull & ND & 75.11 $\pm$ 2.97  & 0.697 $\pm$ 0.014 \\
\hline
\end{tabular}
\end{center}
\end{table}
%

@@

%---------------------------------------------------------------
\section{Conclusions and future work}
\label{sec:conclusion}

@@

In general, linking and finding correlations between user choices and
performance is an interesting avenue to explore in the future.
%  Even if
% the three previous experiments were published in a similar way, one
% obtained up to 5 times more total cycles  than the one with the least
% number of cycles. 
It is also essential to obtain volunteers as
simultaneously as possible, so it is possible that the features of the
social network in terms of real-time use will also play a big
role. Even as it is difficult to create controlled experiments in this
area, it is an interesting challenge to face in the future. % (Paloma) Wrong collocation

This paper also shows that the time needed to find a solution or if it
is eventually found have an influence on the time people devote to
it. It would be so interesting to check the performance for other computationally demanding
tasks, especially ones that cover the middle ground between a problem
finishing in seconds or minutes (Trap) and weeks (Rastrigin F15). 
Besides, the two versions of the framework include as an
algorithmic variant using random population size. We do not think this
has had a big influence on the results and in fact this is not noticed
for F15, which is the last version tested, but it would be interesting to
measure exactly what this influence has been. 
% In general there are
% many issues with the evolutionary algorithm implementation itself,
% including using different, or adaptive, policies for inserting and
% sending individuals to the pool as was done in \cite{araujo2008mam},
% using different policies for population initialization, and also the
% incorporation of high-speed local resources to the pool to check what
% would be the real influence of the volunteer pool to the final
% performance.

@@

% Finally, the implementation needs some refinement in terms of
% programming and also ease of use. Tools such as Yeoman might be used % ¿añadir una cita o la URL http://yeoman.io/ a pie de página ?
% to create a generator in which the user just has to create a fitness
% function, with the rest of the framework wrapped around
% automatically.
%  All this, as well as the data used for this paper and the paper
% itself, has been published with a free license in GitHub at
% \url{https://github.com/JJ/modeling-volunteer-computing}.
